---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(knitr)
```

# MVSE Updates for *Culex pipiens* transmission of West Nile virus

This repository is a fork from [taishiNakase/MVSE] that supplements the manuscript **Jones 2025** focused on customizing the MVSE package to be better suited to predicting West Nile Virus transmission in *Culex pipiens* mosquitos. The fork primarily updated:

-   **mvsemodel.R**

    -   This code was updated to replace RH with VPD, add T to the biting rate equation, and replace *Aedes* priors with *Culex* estimates from literature.

-   **utils.R**

    -   This code was updated to add the Weibull distribution, replace the *Aedes* mortality and biting rate functions with *Culex* versions, and to calculate VPD from RH and T.

The following additional files were added to the /manuscript folder to support development of the paper:

-   **estCulexPriors.Rmd** (and associated html file)

    -   This R Notebook contains the test and diagnostic code used to develop the new functions and priors for the *Culex* version of MVSE.

-   **mvseCustomRuns.Rmd** (and associated html file)

    -   This R Notebook contains the scripts used to run the custom *Culex* MVSE. Due to challenges compiling the package on Mac OS stemming from changes to libraries available in Xcode, this Notebook was run on Ubuntu via a VMware Fusion instance.

-   **mvseOriginalRuns.Rmd** (and associated html file)

    -   This R Notebook contains the scripts used to run the original *Aedes* MVSE. It also contains the code used to prepare the West Nile virus disease data from ArboNet and to obtain the environmental data using GSODR.

# MVSE

MVSE, Mosquito-borne Viral Suitability Estimator, provides methods for the estimation of an index of climate-based transmission potential for mosquito-borne viruses. It can be parameterized for any combination of host, virus and mosquito-species of interest with local climate data.

## Installation

The MVSE package is currently not hosted on CRAN (Comprehensive R Archive Network). We thus need to install the package directly from the GithHub repo where it is hosted.

```{r results='hide', message=FALSE, eval=FALSE}
install.packages("devtools")
install.packages("tidyverse")
require(tidyverse)
require(devtools)

install_github("TaishiNakase/MVSE")
```

## Example

Let's go through a short example of how one might use MVSE to estimate Index P for some mosquito-borne virus.

First, we will install some packages.

```{r message=FALSE}
library(MVSE)
```

We first need to create a `mvse_model` object, which requires time series data for climatic variables (i.e. temperature, humidity and rainfall) as well as epi-entomological prior distributions for the virus/vector/host system of interest.

```{r}
# climate data
data("climateFSA")

# user-defined model
priors <- list(mosq_life_exp=list(pars=c(mean=12, sd=2), dist="normal"),
               mosq_inc_per=list(pars=c(mean=7, sd=2), dist="normal"),
               mosq_biting_freq=list(pars=c(mean=0.25, sd=0.01), dist="normal"),
               human_life_exp=list(pars=c(mean=71.1, sd=2), dist="normal"),
               human_inc_per=list(pars=c(mean=5.8, sd=1), dist="normal"),
               human_inf_per=list(pars=c(mean=5.9, sd=1), dist="normal"))
example_mvse_model <- mvse_model(model_name="user_test", climate_data=climateFSA, priors=priors)
```

Let's take a look at the input data to be used in the sampling procedure.

```{r}
example_mvse_model
```

Next, let's perform the MCMC (Markov chain Monte Carlo) sampling procedure to estimate Index P.

```{r}
example_mvse_fit <- sampling(example_mvse_model, verbose=FALSE)
```

Finally, let's take a look at the estimated distribution of the time series of Index P.

```{r}
indexP_plot <- mcmc_index_dist(example_mvse_fit, index="indexP")
indexP_plot
```
